---
output: 
  github_document: 
    pandoc_args: ["--smart"]
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# lazyTyper

[![Project Status: WIP - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg)](http://www.repostatus.org/#wip)
[![Build Status](https://travis-ci.org/akersting/lazyTyper.svg?branch=develop)](https://travis-ci.org/akersting/lazyTyper)
[![Coverage Status](https://codecov.io/github/akersting/lazyTyper/coverage.svg?branch=develop)](https://codecov.io/github/akersting/lazyTyper)
[![MIT License](https://img.shields.io/badge/license-MIT-brightgreen.svg)](https://opensource.org/licenses/MIT)

lazyTyper adds the concepts of strong and explicit typing to R. Before its first
use, a variable can be declared to be of a specific type. Existing variables can
be casted to a specific type. Assigning values of the wrong type to such typed
variables will throw an error.

In addition to pure type information, it is also possible to associate
additional properties with a variable, e.g. the length of a vector and whether
it may contain NAs or not. Such properties are also validated when the variable
is modified.

## Installation
There is no stable release of lazyTyper yet. To install the latest development
version run:
```{r, eval = FALSE}
devtools::install_github("akersting/lazyTyper@develop")
```

## Introductory Examples

### Declare and Cast
Use `declare` to type a non-existing variable and `cast` to type an existing one.
```{r, error = TRUE}
library(lazyTyper)

a <- 1:3
cast(a, "numeric", length = 2)
cast(a, "numeric", length = 3)
declare(b, "character")

is.typed(a)
is.valid(a)

is.typed(b)
is.valid(b)
```

### Assignment Operators
There are two different operators for typed assignment: `%<-s%` and `%<-%`. The
former one is kind of more "secure", since the variable (apparently) is not
modified if the assignment would violate its type or its properties.
```{r, error = TRUE}
a %<-s% .(1:3)
names(a) %<-% .(c("first", "second", "third"))
a[1] %<-s% .(2)

a[4] %<-s% .(4)
a

class(a) %<-s% .("character")
a
is.valid(a)
```

`%<-%`, on the other hand, might leave the variable in an invalid state.
```{r, error = TRUE}
b %<-% .("Hello World!")

b %<-% .(123)
b
is.valid(b)
```

The advantage of `%<-%` however is that it allows to modify a variable in place,
while `%<-s%` always creates a copy of it. This is due to the laziness of the
type checking: it is only done after the actual assignment was performed. Hence,
`%<-s%` has to create a backup of the variable beforehand, which it can then
restore if the assignment invalidated the variable. The reason behind this 
laziness is that it is not possible to reliably determine how an assignment
will change a variable without actually trying it out. For example, how `var` 
will change due to the following assignment depends on the exact definition of
`myFancyFun<-`.
```{r, eval = FALSE}
myFancyFun(var) <- c(TRUE, NA, FALSE)
```

It is almost always better to use the quicker and more memory friendly `%<-%`,
except maybe when using R interactively.

For both operators the right hand side must be enclosed in `.()`. This ensures
that assignment expressions where the right hand side contains binary operators 
like `+` or `&` are evaluated in correct or --- to be more precise --- expected
order. The necessity for this arises from the high precedence of custom 
operators like `%<-%` and `%<-s%`. Because of this,
```{r, eval = FALSE}
a %<-% 3 + 7
```
would actually be interpreted as
```{r, eval = FALSE}
(a %<-% 3) + 7,
```
which is probably not what we want. Of course, wrapping the right hand side in
normal parentheses would generally solve this problem. However, there is no way
to test for their presence or absence from within the operator functions, while
this can be done for `.()`, which technically is a function call.

#### RStudio Addins
The package ships with addins for RStudio (>= v0.99.1111) for quickly inserting 
`%<-% .()` and `%<-s% .()`. Go to Tools > Addins > Browse Addins... > Keyboard 
Shortcuts... to assign keyboard shortcuts for this.

### Secure Get
For all functions not part of lazyTyper, a typed variable in no way differs from
an untyped one. This also implies that one can use the base assignment operator
to assign an invalid value to a typed variable. Therefore, the function `g` 
offers a secure way to access the value of a type variable.
```{r, error = TRUE}
a <- "I'm supposed to be numeric!"
g(a)
```

For example in tight for-loops one might want to deliberately use the base
assignment operate since it does not perform type checking and hence can be
considerably faster.

### Constants
With `const` it is possible to mark an existing variable as constant. If it is 
modified afterwards, accessing it through `g` will fail.
```{r, error = TRUE}
set.seed(123)
my_const <- runif(5)
const(my_const)
g(my_const)
my_const <- 1:5
g(my_const)
```

## Built-in Types
For now, only type `"any"` and the vector types discussed below are supported.

### Vector Types

The following table shows the supported properties (along the columns) of the currently implemented
built-in vector types (along the rows). They all inherit from / are aliases of the hidden (base) type
`.vector`. 

A cell which is not crossed out indicates that the respective property can be set when using that type with `declare` or `cast`. The value in such a cell is the default value of the property for that type. 

A crossed out cell indicates that the respective property cannot be set for that type, i.e. it is fixed to the (crossed out) value in the cell. 

The meaning of a completely empty or crossed out empty cell, i.e of an unset property, is given by the first row.

|          Type |     length    |   min_length  |   max_length  |      set      |      min      |      max      |     whole     |    pattern    | allow_duplicates |    allow_NA   |   allow_NaN   | allow_NULL | allow_missing |       type      |
|--------------:|:-------------:|:-------------:|:-------------:|:-------------:|:-------------:|:-------------:|:-------------:|:-------------:|:----------------:|:-------------:|:-------------:|:----------:|:-------------:|:---------------:|
|   **.vector** | _not checked_ | _not checked_ | _not checked_ | _not checked_ | _not checked_ | _not checked_ | _not checked_ | _not checked_ |   _not checked_  | _not checked_ | _not checked_ | _no/FALSE_ |   _no/FALSE_  |      _N/A_      |
|    **vector** |               |               |               |    &mdash;    |    &mdash;    |    &mdash;    |    &mdash;    |    &mdash;    |                  |    &mdash;    |    &mdash;    |            |               |   ~~"vector"~~  |
|   **logical** |               |               |               |    &mdash;    |    &mdash;    |    &mdash;    |    &mdash;    |    &mdash;    |      &mdash;     |               |    &mdash;    |            |               |  ~~"logical"~~  |
|      **bool** |     ~~1~~     |    &mdash;    |    &mdash;    |    &mdash;    |    &mdash;    |    &mdash;    |    &mdash;    |    &mdash;    |      &mdash;     |     FALSE     |    &mdash;    |            |               |  ~~"logical"~~  |
|   **numeric** |               |               |               |               |               |               |               |    &mdash;    |                  |               |               |            |               |  ~~"numeric"~~  |
|   **integer** |               |               |               |    &mdash;    |               |               |    ~~TRUE~~   |    &mdash;    |                  |               |               |            |               |  ~~"numeric"~~  |
|     **count** |               |               |               |    &mdash;    |     ~~0~~     |               |    ~~TRUE~~   |    &mdash;    |                  |               |               |            |               |  ~~"numeric"~~  |
|    **scalar** |     ~~1~~     |    &mdash;    |    &mdash;    |               |               |               |               |    &mdash;    |      &mdash;     |     FALSE     |     FALSE     |            |               |  ~~"numeric"~~  |
| **character** |               |               |               |               |    &mdash;    |    &mdash;    |    &mdash;    |               |                  |               |    &mdash;    |            |               | ~~"character"~~ |
|    **string** |     ~~1~~     |    &mdash;    |    &mdash;    |               |    &mdash;    |    &mdash;    |    &mdash;    |               |      &mdash;     |     FALSE     |    &mdash;    |            |               | ~~"character"~~ |
#### The Properties
- `length`/`min_length`/`max_length`: the allowed exact/minimum/maximum length of the vector. If `length` is set then `min_length` and `max_length` cannot be set. `max_length` must be larger equal `min_length`.
- `set`: a vector of allowed values. If set, then `min`, `max`, `whole` and `pattern` cannot be set. If `set` contains at least one `NA`/`NaN` then `allow_NA`/`allow_NaN` is set to `TRUE`.
- `min`/`max`:  the allowed minimum/maximum value to be stored in the vector. `max` must be larger equal `min`.
- `whole`: if `TRUE` only approximately whole numbers may be stored in the vector, i.e. the difference between all values and their nearest whole number must be less or equal `.Machine$double.eps^0.5`.
- `pattern`: only allow values in the vector which this regular expression matches.
- `allow_duplicates`: are duplicated values allowed in the vector? `NA` and `NaN` (for numeric vectors) are treated as incomparable.
- `allow_NA`/`allow_NaN`: are `NA`/`NaN` values allowed in the vector? `allow_NA` here strictly refers only to `NA` (and not `NaN`), i.e. `any(is.na(x) & !is.nan(x))` is used to test this. If `set` contains `NA`/`NaN` then `allow_NA`/`allow_NaN` must not be `FALSE`.
- `allow_NULL`: may the whole variable be `NULL`? Note that the default does not allow this, i.e. not setting `allow_NULL` it is equivalent to setting `allow_NULL = FALSE`.
- `allow_missing`: may the variable be a missing argument, i.e. identical to the empty name? This is relevant when casting a formal argument of a function. Note that the default does not allow this, i.e. not setting `allow_missing` it is equivalent to setting `allow_missing = FALSE`.

## Registering Custom Types
ToDo: explain how to register custom types and overload built-in types.


## Advanced: Dynamic Properties
In `declare` and `cast` the additional properties can be language objects (names,
calls or expressions). Every time the typed variable is checked, such property
specifications are re-evaluated in the environment of the variable:

```{r}
X <- iris[1:100, ]
declare("weights", "numeric", length = quote(nrow(X)))
weights %<-% .(rep(1, 100))

X <- iris[1:50, ]
is.valid(weights)
```


