---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# lazyTyper

[![Build Status](https://travis-ci.org/akersting/lazyTyper.svg?branch=develop)](https://travis-ci.org/akersting/lazyTyper)
[![Coverage Status](https://codecov.io/github/akersting/lazyTyper/coverage.svg?branch=develop)](https://codecov.io/github/akersting/lazyTyper)

lazyTyper adds the concept of strong typing to R. Before its first use, a
variable can be declared to be of a specific type. Existing variables can be 
casted to a specific type. Assigning values of the wrong type to such typed
variables will throw an error.

In addition to pure type information, it is also possible to associate
additional properties with a variable, e.g. the length of a vector. Such
properties are also validated when the variable is modified.

## Installation
There is no stable release of lazyTyper yet. To install the latest development
version run:
```{r, eval = FALSE}
devtools::install_github("akersting/lazyTyper@develop")
```

## Introductory Examples

### Declare and Cast
Use `declare` to type a non-existing variable and `cast` to type an existing one.
```{r, error = TRUE}
library(lazyTyper)

a <- 1:3
cast(a, "numeric", length = 2)
cast(a, "numeric", length = 3)
declare(b, "character")

is.typed(a)
is.valid(a)

is.typed(b)
is.valid(b)
```

### Assignment Operators
There are two different operators for typed assignment: `%<-s%` and `%<-%`. The
former one is kind of more "secure", since the variable (apparently) is not
modified if the assignment would violate its type or its properties.
```{r, error = TRUE}
a %<-s% 1:3
names(a) %<-% c("first", "second", "third")
a[1] %<-s% 2

a[4] %<-s% 4
a

class(a) %<-s% "character"
a
is.valid(a)
```

`%<-%`, on the other hand, might leave the variable in an invalid state.
```{r, error = TRUE}
b %<-% "Hello World!"

b %<-% 123
b
is.valid(b)
```

The advantage of `%<-%` however is that it allows to modify a variable in place,
while `%<-s%` always creates a copy of it. This is due to the laziness of the
type checking: it is only done after the actual assignment was performed. Hence,
`%<-s%` has to create a backup of the variable beforehand, which it can then
restore if the assignment invalidated the variable. The reason behind this 
laziness is that it is not possible to reliably determine how an assignment
will change a variable without actually trying it out. For example, how `var` 
will change due to the following assignment depends on the exact definition of
`myFancyFun<-`.
```{r, eval = FALSE}
myFancyFun(var) <- c(TRUE, NA, FALSE)
```

It is almost always better to use the quicker and more memory friendly `%<-%`,
except maybe when using R interactively.

#### RStudio Addins
The package ships with addins for RStudio (>= v0.99.878) for quickly inserting 
"%<-%" and "%<-s%". Go to Tools > Addins > Browse Addins... > Keyboard 
Shortcuts... to assign keyboard shortcuts for this.

### Secure Get
For all functions not part of lazyTyper, a typed variable in no way differs from
an untyped one. This also implies that one can use the base assignment operator
to assign an invalid value to a typed variable. Therefore, the function `g` 
offers a secure way to access the value of a type variable.
```{r, error = TRUE}
a <- "I'm supposed to be numeric!"
g(a)
```

For example in tight for-loops one might want to deliberately use the base
assignment operate since it does not perform type checking and hence can be
considerably faster.

### Constants
With `const` it is possible to mark an existing variable as constant. If it is 
modified afterwards, accessing it through `g` will fail.
```{r, error = TRUE}
my_const <- runif(5)
const(my_const)
g(my_const)
my_const <- 1:5
g(my_const)
```

## Built-in Types
Currently only "any", "numeric" and "character" are supported, the latter two 
only partially.

ToDo: describe the types with all their properties.


### Additional Properties, Reference Counting and Copying
ToDo: detail why checking certain properties will lead to a copy the next time
the variable is modified.

## Registering Custom Types
You can easily register a custom type. To do so you need three things:

* a type name, e.g. "symmetric_matrix". By using the name of a built-in type
you can overload it *for all newly typed variables* -- already typed variables
will keep their built-in type. 
* a function to check the validity of the additional properties passed to either
declare or cast when creating a typed variable. This function must accept these
additional properties as (named) parameters and it should throw an error if 
properties used are invalid. Afterwards, it must return -- as a (named) list -- 
the set of properties which should finally be stored. This set can be different
from the original arguments to this function, e.g. because one of two 
inconsistent properties was removed. There is the helper function `args2list`
in the package, which returns the *current* values of the arguments of the 
calling function as a list.
* a quoted expression used to test the validity of a typed variable by e.g.
`is.valid`, `%<-%` or `g`. This expression is evaluated in the environment of 
the variable to test and should hence (ideally) not create any (temporary) 
objects. Within this expression you must refer to the variable to test by 
`.XXX.` and the list of properties is available as `.lazyTyper_properties.` 
This expression should never fail but rather set the logical value 
`.lazyTyper_valid` to `FALSE` if the variable is invalid. In this case the 
"error"-attribute of the variable `.lazyTyper_valid` should contain the 
reason(s) for the invalidity as a character vector. The recommended way to set 
`lazyTyper_valid` to `FALSE` and to add a(nother) error message to it is by 
calling the helper function 
`setInvalidWError("This is pasted to ", "a single error message!")`.

ToDo: explain how to register custom types and overload built-in types.
